A {practical} review on linear and nonlinear global approaches to stability analysis

### 2018-06-15 ARTICLE_STABFEM/ARTICLE_ASME_Submitted.tex
Documentation !

The stability of fluid flows usually depends on the value of a given parameter. A bifurcation occurs when a critical value is reached and the original solution becomes linearly unstable, with the system tending towards a new steady or unsteady state. 
%The destabilization of a flow as a parameter is varied, leading from a steady solution to a unsteady one, is an ubiquous situation in fluid dynamics. 
In the second part of the 19th century, specific analytical and numerical methods able to study these bifurcations have emerged and are continuously evolving up to the present days. 
A crucial point, that drove the development in this field, is the availability of larger and larger computing resources. 
Initially, the linear stability theory focused on fluid flows that are homogeneous in two spatial directions, e.g. plane Poiseuille flow \cite{Dreid2004}. This implies that the streamwise and the spanwise base-flow gradients vanish and, as a consequence, it is possible to consider only one (streamwise) velocity component. Thus, the stability problem requires a \textit{local} numerical resolution, i.e. a one-dimensional problem.
On the other hand,  when there are at least two spatial variables, the class of methods suited to solve such problems are generally called "global stability approaches".
%When the flow has no analytical expression, the stability problem requires a numerical resolution, and when there are at least two spatial variables, the class of methods suited to solve such problems are generally called "global stability approaches."%\footnote{Sorry Laurette !}
A classical example of such behavior in fluid dynamics is the instability occurring in the wake of a circular cylinder. At low Reynolds number (precisely for $Re < 46.7$) the flow is steady and symmetric, but for larger values of Re a global instability arises in the flow field leading to the well-known von K\'arm\'an vortex alley. 
%Among all instability problems, one of the most famous and most studied is the flow around a 2D cylinder, which becomes unsteady for $Re \approx 47$ leading to the well-known Bénard-Von Karman vortex alley. 
This flow configuration has served as a benchmark in the development of this class of methods.  
If one is only interested in predicting the stability or instability of a flow, it is enough to conduct a {\em linear stability analysis} which is the fundamental brick of global stability approaches. 
Beyond this simple question, in the past two decades, a number of extensions have been developed and popularized.  {\em Adjoint methods} are an important extension \cite{GiannettiLuchini},\cite{Marquet} ; they can give insight into the sensitivity of the flow to intrinsic or extrinsic contributions. 
{\em Nonlinear stability approaches} \cite{MLugo2014}, \cite{SippLebedev} have also been developed in order to extend the range of applicability of the approach towards large amplitude perturbations.

@@

The objective of the present work is to contribute to the popularization of such methods 
in two ways:
%for both research and education purposes. 
\begin{itemize}
\item
First, we give a concise but self-contained exposition of the main concepts and 
specific numerical methods pertaining to global stability, including basic linear stability, adjoint-based sensitivity, as well as the most recent nonlinear developments.
\item
Secondly we offer an open-source and user-friendly software called  "StabFem"
\cprotect\footnote{The StabFem software may be obtained at the following url : 
\url{https://github.com/erbafdavid/StabFem}} 
to perform such calculations. The software combines program written in both FreeFem++ and Matlab languages. 
FreeFem++ is used to generate and adapt the meshes and to solve the various linear problems arising in the analysis. Matlab is used as a driver to monitor the computations, perform the required loops over parameters, and plot the results.
\end{itemize}

@@

Although we don't claim to invent any radically new method, our exposition and implementation contains a number of originalities making the computation particularly efficient in terms of computational time and memory (all the figures of the paper can be produced in only a few minutes on a standard laptop)
\footnote{All the figures contained in the present paper can be processed by launching a single Matlab script SCRIPT\_CYLINDER\_ALLFIGURES.m . On a MacbookPro (2018,  2.5Ghz, 16Go Ram),  the execution time is 19.76 seconds for the results of the linear analysis (section 3) and 15. seconds for the results of the nonlinear analysis (section 4).
}. 
The most notable originalities are the systematic use of mesh adaptation (\S 2 and 3), the use of simple shift-invert instead of Arnoldi (\S 3), and a reformulation and simplification of the nonlinear self-consistent approach of Mantic-Lugo et al  (\S 4).

@@

We start from the general problem of a flow field $[{\bf u},p]$ satisfying the incompressible Navier-Stokes equations on a domain $\Omega$,
\begin{eqnarray} \label{NSprimitive}
\partial_t {\bf u} = {\color{red}{\cal NS} ({\bf u},p)}
\equiv - {\bf u} \cdot \nabla {\bf u} - \nabla p + \frac{2}{Re}  \nabla \cdot {\mathsf{D}}({\bf u}),  \\
\nabla \cdot {\bf u} = 0,
\end{eqnarray}
with suitable boundary conditions on the frontier $\partial \Omega$ of the domain.
Here $ {\mathsf{D}}({\bf u}) $ is the rate-of-strain tensor defined as
$$
 {\mathsf{D}}({\bf u}) = 1/2
\left( \nabla {\bf u} +  \nabla^T  {\bf u}\right),
$$ 
In the framework of finite element methods, we need to write the equation in weak form.
Prior to this we define a scalar product as follows, for either scalar or vectorial quantities 
$\left< \phi_1, \phi_2 \right> $:
$$
\left< \phi_1, \phi_2 \right> = \int_\Omega \overline{\phi}_1 \cdot \phi_2   \mbox{ d} \Omega,
$$
The weak form of the Navier-Stokes equations is readily defined by introducing test functions 
$[{\bf v},q]$ associated with the momentum and continuity equations, and integrating over the domain\footnote{In the simple presentation given here we have omitted the issue of boundary conditions. Details on way boundary conditions can be incorporated in the weak formulation through integration by parts can be found in appendix A.}
\be{NSweak}
\forall [{\bf v},q], \quad \partial_t \left< {\bf v}, {\bf u}\right> = \left< {\bf v} , {\color{red}{\cal NS} ({\bf u},p)} \right> + \left< q, \nabla \cdot {\bf u}\right>.
\ee

@@

We look for a steady base-flow {\color{red}$[{\bf u}_b,p_b]$} satisfying the steady Navier-Stokes equations, i.e. 
%${\cal NS} ({\bf u}_b,p_b) = 0$.
{\color{red}${\cal NS} ({\bf u}_b,p_b) = 0$.}
Suppose that we have a {\color{red}``guess''}  for the base flow $[{\bf u}_b^g,p_b^g]$  which almost satisfies the equations.  We look for a better approximation under the form
\be{Newton1}
[{\bf u}_b,p_b]  = [{\bf u}_b^g,p_b^g] + [\delta {\bf u}_b, \delta p_b].
\ee
Injecting (\ref{Newton1}) into the weak form (\ref{NSweak}) of the Navier-Stokes equations and developing up to linear terms in terms of the perturbation lead to  
${\cal NS}  ({\bf u}_b^g,p_b^g) + {\color{red} {\cal LNS} _{{\bf u}_b^g}(\delta {\bf u}_b,\delta p_b)}$, which can also be written in weak form as :
\bes{Newton2}
&\left< {\bf v}, {\color{red}{\cal NS} ({\bf u}_b^g,p_b^g)}\right> + \left<q, \nabla \cdot {\bf u}_b^g\right>  
\\
+ &\left< { \bf v}, {\color{red}{\cal LNS} _{{\bf u}_b^g}( \delta {\bf u}_b,\delta p_b) }\right> + \left<q, \nabla \cdot \delta{\bf u}_b\right> = 0,
\ees
where ${\cal LNS}$ is the linearised Navier-Stokes operator, defined by its action on a flow field {\color{red}$[{\bf u} , p]$} as follows 
\be{defNSL}
{\color{red} {\cal LNS}_{{\bf U}}( {\bf u}, p)} = - {\cal C}( {\bf U} , {\bf u}) -\nabla p
+\frac{2}{Re} \nabla  \cdot {\mathsf{D}}({\bf u}), %\quad ( \mbox{ with } \nabla \cdot {\bf u } = 0 \, ).
 \ee
and ${\cal C}$ is the convection operator defined by 
\be{defC}
{\cal C}( {\bf U} , {\bf u}) = \left( {\bf U} \cdot \nabla \right) {\bf u} + \left( {\bf u} \cdot \nabla \right)  {\bf U}.
\ee
This problem can now be discretized by projecting upon a basis of Taylor-Hood $(u,v,p) \rightarrow (P2,P2,P1)$ finite elements. Noting $\delta X$ the discretization of {\color{red}$[\delta {\bf u}_b,\delta p_b]$} this eventually leads to a matricial problem of the form $A \cdot \delta X = Y$. The procedure of Newton iteration is to solve iteratively this set of equations up to convergence.
In our implementation, the algorithm is written in the Freefem++ solver {\em Newton\_2D.edp} 
which is wrapped by the Matlab driver {\em Freefem\_BaseFlow.m}.

@@

\paragraph{Direct eigenvalue problem}
We study the onset of the instability within the linear theory by using a normal-mode analysis:
\be{startmodepropre}
{\bf u} ({\bf{x}},t) = {\bf u}_b({\bf{x}}) + \epsilon \hat{\bf u}({\bf{x}}) e^{\lambda t}, \,\,\,\, {p}({\bf{x}},t) = {p}_b({\bf{x}}) + \epsilon \hat{p}({\bf{x}}) e^{\lambda t},
\ee
where $\lambda = \sigma + i \omega$ is the eigenvalue, $\sigma$ the amplification rate,
$\omega$ the oscillation rate, $\hat{{\bf u}},\hat{p}$ the eigenmodes, and $\epsilon$ a small parameter.
The eigenmodes and the eigenvalues are the solution of the following eigenproblem:
 \be{LEP}
\lambda \hat{{\bf u}} = {\cal LNS}_{{\bf u}_b}( \hat{\bf u},\hat{p}),
\ee
or, in weak form : 
\be{eigenvalueproblem}
\lambda \left< {\bf v} , \hat{{\bf u}} \right> = \left< {\bf v}, {\cal LNS}_{{\bf u}_b} ( \hat{\bf u},\hat{p})\right> + \left< q, \nabla \cdot \hat{\bf u} \right>.
\ee
After discretization, we end up with an eigenvalue problem with the matricial form
\be{Eigen_matricial}
\lambda {\mathsf{B}} \hat{X} = {\mathsf{A}} \hat{X},
\ee
{\color{red} where} ${\mathsf{A}}$ is the matrix resulting from the discretization of ${\cal LNS}_{{\bf u}_b}$, i.e. the same matrix  appearing in the Newton computation of the base flow, and  $B$ is a 'weight' matrix associated to the scalar product $\left<v,u\right> = \int \overline{\bf v} \cdot {\bf u} \mbox{ d} \Omega$.

@@

When it comes to the numerical resolution of generalized eigenvalue problems such as $A X = \lambda B X$ (or its adjoint version (\ref{Eigen_Adj_matricial}) ), several methods are possible. Direct methods to compute the whole spectrum are both costly prohibitive and useless. A popular alternative is the use of iterative methods which allow us to compute a limited set of eigenvalues located in the vicinity of a 
"shift" value $\lambda_{shift}$. The simplest version of this method is the simple shift-invert iteration, which consists of solving iteratively the system
$$
X^{n} =  (A- \lambda_{shift} B)^{-1} B X^{n-1}.
$$ 
It is easy to show that this iterative procedure quickly asymptotes to $X^{n+1} \approx ({\lambda^*}^{-1})^n \hat{X}$
where $\hat{X}$ is the eigenmode with largest ${\lambda^*}^{-1}$ (i.e. the one with eigenvalue $\lambda$ closest to the shift).

When a good estimation of the eigenvalue is available, this method converges very rapidly and is very efficient, but it can only provide a single eigenvalue.
If we want to compute a larger number of eigenvalues, we can revert to a generalized version of iterative methods, called Arnoldi methods\cite{Arnoldi51}. The shift-invert version of the Arnoldi method is in fact the most commonly used method of the current time and is at the basis of both the popular matlab function {\tt{eigs}} and the eigenvalue solver of FreeFem (i.e. ARPACK++). Our implementation in StabFem allows to chose between single eigenvalue computation (power method) and multiple eigenvalue computation (Arnoldi). The selection is made according to the parameter "nev" transmitted to the driver.

@@

As for any numerical method, a crucial point in the numerical efficiency is the design of the mesh. The finite element method allows to use unstructured mesh and hence to locally adapt the refinement. %Following Sipp \& Lebedev, 
The most common procedure is to decompose the domain into several parts with different grid densities; for instance for the wake of a cylinder, we will design a near-wall region with very small  size, a "wake" region with intermediate mesh size, and an outer region with large mesh size. The inconvenient is that the design relies on an a priori expectation of the regions where gradients will be large.

@@

Let's consider now the matlab code reported in figure \ref{Listing2}. First we build an initial mesh (line 1), and compute base flow solutions for increasing values of the Reynolds number up to $Re = 60$ (lines 3-9).
Then we perform the mesh adaptation with the structural sensitivity, as explained in a previously (line 13). 
The resulting mesh, depicted in figure \ref{meshM2}, is used for the rest of the computations presented in this paper (except for displaying the eigenmode in figure 4$a$). 
Appendix D presents additional test regarding mesh convergence, and demonstrate that results obtained with the resulting mesh are trustable within $0.1\%$ accuracy for the eigenvalue, while requiring a very reasonable number of grid points compared to previous studies of this problem.

@@

In the past decade, efforts have been devoted to extend the range of validity of global approaches into the nonlinear regime for $Re>Re_c$, with the double objective to describe the properties of the limit cycle reached after saturation and to derive amplitude equations describing the transitent dynamics towards this cycle. The two main milestones in this direction {\color{red} are} the {\em weakly nonlinear model}  (WNL) of Sipp \& Lebedev \cite{SippLebedev} and the {\em self-consistent model} (SC) of Manti\v{c}-Lugo et al \cite{MLugo2014}.
 {\color{blue} In this review, we will address only the first of the two questions, namely {\color{red} the} description of the saturated cycle, and leave aside the question of transient dynamics. We will successively review the two aforementioned models, in a simplified formulation devoted to describe only the saturated cycle. We also provide a simple implementation of these two models. Finally, in line with the previous section, we will again illustrate with results obtained for the wake of a cylinder.
 }

@@

 In this part, to simplify the notations, we symbolically write the Navier-Stokes equations, as {\color{red}$\partial_t {\bf u} = {\cal NS} ({\bf u})$}, therefore dropping the systematic reference to the incompressibility constraint and associated pressure field. {\color{red} The same is done with the linearized operator ${\cal LNS}_{{\bf U}}( {\bf u})$.}

@@

\paragraph{WNL model : analysis}

@@

We first review the weakly nonlinear model of Sipp \& 
Lebedev\cite{SippLebedev}, also discussed by Gallaire et al. \cite{FDR2016}.  
The initial derivation of \cite{SippLebedev} makes use of a multiple scale method in order to obtained an amplitude equation.
This complete analysis is reproduced in appendix B. In the present paragraph, we give a simplified derivation of this model restricted to the description of the periodic saturated cycle.
The starting point can be taken as the following expansion of the velocity flow field:
 {\color{red}
\be{WNL1}
\begin{aligned}
&{\bf u} = {}  {\bf u}_{bc} + \epsilon \left[ A_{wnl}  \hat{\bf u} e^{i (\omega_c+\epsilon^{\color{red}2} \omega_\epsilon)  t} + c.c. \right]\\
&+ \epsilon^2 \left[ {\bf u}_\epsilon + |A_{wnl}|^2  {\bf u}_{2,0} + \left(  A_{wnl}^2 {\bf u}_{2,2} e^{2 i(\omega_c+\epsilon^{\color{red}2} \omega_\epsilon)  t} + c.c. \right) \right]\\
& + {\cal O}( \epsilon ^3),
\end{aligned}
\ee
}

This expansion is built as an asymptotic expansion in terms of the small parameter $\epsilon = \sqrt{1/Re_c - 1/Re}$ corresponding to the distance to the critical Reynolds. 
The zero-order term {\color{red}${\bf u}_{bc}$} is the base flow at the threshold $Re_c$.

@@

The second-order term contains three contributions: ${\bf u}_{\epsilon}$ is the modification to the base flow related to the 
increase of $Re$, ${\bf u}_{2,0}$  represents the nonlinear interaction of $\hat{\bf u}$ with its conjugate, ${\bf u}_{2,2}$ the nonlinear 
interaction of $\hat{\bf u}$ with itself. These three terms are computed as solutions of nonsingular linear systems 
(see eqs. \ref{eq:LL1},\ref{eq:LL2},\ref{eq:LL3} in appendix B).

At ${\cal O}( \epsilon ^3)$, compatibility conditions have to be imposed to ensure that the problem is correcty posed. These conditions lead to an "amplitude equation" which relates the amplitude $ A_{wnl}$ to three parameters $\Lambda, \nu_{0}$ and $\nu_{2}$ which depend
uniquely on ${\bf u}_\epsilon$, ${\bf u}_{2,0}$ and ${\bf u}_{2,2}$, respectively (see equations \ref{LAMBDA}, \ref{NU0} and \ref{NU2} in appendix B). Restricting to the description of the limit cycle, the "amplitude equation" takes the form :

@@

Finally, as specified above, we explain how the mean drag $F_{x,0}$ and fundamental components $(F_{y,1,c},F_{y,1,s})$ of the oscillating lift can be predicted from the WNL approach.
The mean drag  can be obtained by $F_{x,0} = {\cal D}_{Re}({\bf u}_m,p_m)$, where ${\cal D}$ is the drag operator defined in \ref{drag} and 
$[{\bf u}_m,p_m]$ is the {\em mean flow}  
which corresponds to the time-average of expansion  \ref{WNL1}, namely :
$
[{\bf u}_m,p_m] = [{\bf u}_{bc},p_{bc}] +\epsilon^2 \left( [{\bf u}_\epsilon,p_\epsilon] + A_{wnl}^2 [{\bf u}_{2,0}, p_{2,0} ] \right).
$
Developing the terms as an asymptotic expansion leads to :

>>>>

The stability of fluid flows usually depends on the value of a given parameter. A bifurcation occurs when a critical value is reached and the original solution becomes linearly unstable, with the system tending towards a new steady or unsteady state. 
%The destabilization of a flow as a parameter is varied, leading from a steady solution to a unsteady one, is an ubiquous situation in fluid dynamics. 
In the second part of the 19th century, specific analytical and numerical methods able to study these bifurcations have emerged and are continuously evolving up to the present days. 
A crucial point, that drove the development in this field, is the availability of larger and larger computing resources. 
Initially, the linear stability theory focused on fluid flows that are homogeneous in two spatial directions, e.g. plane Poiseuille flow \cite{Dreid2004}. This implies that the streamwise and the spanwise base-flow gradients vanish and, as a consequence, it is possible to consider only one (streamwise) velocity component. Thus, the stability problem requires a \textit{local} numerical resolution, i.e. a one-dimensional problem.
On the other hand,  when there are at least two spatial variables, the class of methods suited to solve such problems are generally called ``global stability approaches".
%When the flow has no analytical expression, the stability problem requires a numerical resolution, and when there are at least two spatial variables, the class of methods suited to solve such problems are generally called "global stability approaches."%\footnote{Sorry Laurette !}
A classical example of such behavior in fluid dynamics is the instability occurring in the wake of a circular cylinder. At low Reynolds number (precisely for $Re < 46.7$) the flow is steady and symmetric, but for larger values of Re a global instability arises in the flow field leading to the well-known von K\'arm\'an vortex alley. 
%Among all instability problems, one of the most famous and most studied is the flow around a 2D cylinder, which becomes unsteady for $Re \approx 47$ leading to the well-known Bénard-Von Karman vortex alley. 
This flow configuration has served as a benchmark in the development of this class of methods.  
If one is only interested in predicting the stability or instability of a flow, it is enough to conduct a {\em linear stability analysis} which is the fundamental brick of global stability approaches. 
Beyond this simple question, in the past two decades, a number of extensions have been developed and popularized.  {\em Adjoint methods} are an important extension \cite{GiannettiLuchini},\cite{Marquet}; they can give insight into the sensitivity of the flow to intrinsic or extrinsic contributions. 
{\em Nonlinear stability approaches} \cite{MLugo2014}, \cite{SippLebedev} have also been developed in order to extend the range of applicability of the approach towards large amplitude perturbations.

@@

The objective of the present work is to contribute to the popularization of such methods 
in two ways:
%for both research and education purposes. 
\begin{itemize}
\item
First, we give a concise but self-contained exposition of the main concepts and 
specific numerical methods pertaining to global stability, including basic linear stability, adjoint-based sensitivity, as well as the most recent nonlinear developments.
\item
Secondly we offer an open-source and user-friendly software called  ``StabFem"\cprotect\footnote{The StabFem software may be obtained at the following url : 
\url{https://github.com/erbafdavid/StabFem}} 
to perform such calculations. The software combines program written in both FreeFem++ and Matlab languages. 
FreeFem++ is used to generate and adapt the meshes and to solve the various linear problems arising in the analysis. Matlab is used as a driver to monitor the computations, perform the required loops over parameters, and plot the results.
\end{itemize}

@@

Although we don't claim to invent any radically new method, our exposition and implementation contains a number of originalities making the computation particularly efficient in terms of computational time and memory (all the figures of the paper can be produced in only a few minutes on a standard laptop)\footnote{All the figures contained in the present paper can be processed by launching a single Matlab script SCRIPT\_CYLINDER\_ALLFIGURES.m . On a MacbookPro (2018,  2.5Ghz, 16Go Ram),  the execution time is 19.76 seconds for the results of the linear analysis (section 3) and 15. seconds for the results of the nonlinear analysis (section 4).
}. 
The most notable originalities are the systematic use of mesh adaptation (\S 2 and 3), the use of simple shift-invert instead of Arnoldi (\S 3), and a reformulation and simplification of the nonlinear self-consistent approach of Mantic-Lugo et al  (\S 4).

@@

We start from the general problem of a flow field $[{\bf u},p]$ satisfying the incompressible Navier-Stokes equations on a domain $\Omega$,
\begin{eqnarray} \label{NSprimitive}
\partial_t {\bf u} = {\cal NS} ({\bf u},p)
\equiv - {\bf u} \cdot \nabla {\bf u} - \nabla p + \frac{2}{Re}  \nabla \cdot {\mathsf{D}}({\bf u}),  \\
\nabla \cdot {\bf u} = 0,
\end{eqnarray}
with suitable boundary conditions on the frontier $\partial \Omega$ of the domain.
Here $ {\mathsf{D}}({\bf u}) $ is the rate-of-strain tensor defined as
$$
 {\mathsf{D}}({\bf u}) = 1/2
\left( \nabla {\bf u} +  \nabla^T  {\bf u}\right),
$$ 
In the framework of finite element methods, we need to write the equation in weak form.
Prior to this we define a scalar product as follows, for either scalar or vectorial quantities 
$\left< \phi_1, \phi_2 \right> $:
$$
\left< \phi_1, \phi_2 \right> = \int_\Omega \overline{\phi}_1 \cdot \phi_2   \mbox{ d} \Omega,
$$
The weak form of the Navier-Stokes equations is readily defined by introducing test functions 
$[{\bf v},q]$ associated with the momentum and continuity equations, and integrating over the domain\footnote{In the simple presentation given here we have omitted the issue of boundary conditions. Details on way boundary conditions can be incorporated in the weak formulation through integration by parts can be found in appendix A.}
\be{NSweak}
\forall [{\bf v},q], \quad \partial_t \left< {\bf v}, {\bf u}\right> = \left< {\bf v} , {\cal NS} ({\bf u},p) \right> + \left< q, \nabla \cdot {\bf u}\right>.
\ee

@@

We look for a steady base-flow $[{\bf u}_b,p_b]$ satisfying the steady Navier-Stokes equations, i.e. 
%${\cal NS} ({\bf u}_b,p_b) = 0$.
${\cal NS} ({\bf u}_b,p_b) = 0$.
Suppose that we have a ``guess''  for the base flow $[{\bf u}_b^g,p_b^g]$  which almost satisfies the equations.  We look for a better approximation under the form
\be{Newton1}
[{\bf u}_b,p_b]  = [{\bf u}_b^g,p_b^g] + [\delta {\bf u}_b, \delta p_b].
\ee
Injecting (\ref{Newton1}) into the weak form (\ref{NSweak}) of the Navier-Stokes equations and developing up to linear terms in terms of the perturbation lead to  
${\cal NS}  ({\bf u}_b^g,p_b^g) +  {\cal LNS} _{{\bf u}_b^g}(\delta {\bf u}_b,\delta p_b)$, which can also be written in weak form as :
\bes{Newton2}
&\left< {\bf v}, {\cal NS} ({\bf u}_b^g,p_b^g)\right> + \left<q, \nabla \cdot {\bf u}_b^g\right>  
\\
+ &\left< { \bf v}, {\cal LNS} _{{\bf u}_b^g}( \delta {\bf u}_b,\delta p_b) \right> + \left<q, \nabla \cdot \delta{\bf u}_b\right> = 0,
\ees
where ${\cal LNS}$ is the linearised Navier-Stokes operator, defined by its action on a flow field $[{\bf u} , p]$ as follows 
\be{defNSL}
 {\cal LNS}_{{\bf U}}( {\bf u}, p) = - {\cal C}( {\bf U} , {\bf u}) -\nabla p
+\frac{2}{Re} \nabla  \cdot {\mathsf{D}}({\bf u}), %\quad ( \mbox{ with } \nabla \cdot {\bf u } = 0 \, ).
 \ee
and ${\cal C}$ is the convection operator defined by 
\be{defC}
{\cal C}( {\bf U} , {\bf u}) = \left( {\bf U} \cdot \nabla \right) {\bf u} + \left( {\bf u} \cdot \nabla \right)  {\bf U}.
\ee
This problem can now be discretized by projecting upon a basis of Taylor-Hood $(u,v,p) \rightarrow (P2,P2,P1)$ finite elements. Noting $\delta X$ the discretization of $[\delta {\bf u}_b,\delta p_b]$ this eventually leads to a matricial problem of the form $A \cdot \delta X = Y$. The procedure of Newton iteration is to solve iteratively this set of equations up to convergence.
In our implementation, the algorithm is written in the Freefem++ solver {\em Newton\_2D.edp} 
which is wrapped by the Matlab driver {\em Freefem\_BaseFlow.m}.

@@

\paragraph{Direct eigenvalue problem}
We study the onset of the instability within the linear theory by using a normal-mode analysis:
\be{startmodepropre}
{\bf u} ({\bf{x}},t) = {\bf u}_b({\bf{x}}) + \epsilon \hat{\bf u}({\bf{x}}) e^{\lambda t}, \,\,\,\, {p}({\bf{x}},t) = {p}_b({\bf{x}}) + \epsilon \hat{p}({\bf{x}}) e^{\lambda t},
\ee
where $\lambda = \sigma + i \omega$ is the eigenvalue, $\sigma$ the amplification rate,
$\omega$ the oscillation rate, $\hat{{\bf u}},\hat{p}$ the eigenmodes, and $\epsilon$ a small parameter.
The eigenmodes and the eigenvalues are the solution of the following eigenproblem:
 \be{LEP}
\lambda \hat{{\bf u}} = {\cal LNS}_{{\bf u}_b}( \hat{\bf u},\hat{p}),
\ee
or, in weak form : 
\be{eigenvalueproblem}
\lambda \left< {\bf v} , \hat{{\bf u}} \right> = \left< {\bf v}, {\cal LNS}_{{\bf u}_b} ( \hat{\bf u},\hat{p})\right> + \left< q, \nabla \cdot \hat{\bf u} \right>.
\ee
After discretization, we end up with an eigenvalue problem with the matricial form
\be{Eigen_matricial}
\lambda {\mathsf{B}} \hat{X} = {\mathsf{A}} \hat{X},
\ee
where ${\mathsf{A}}$ is the matrix resulting from the discretization of ${\cal LNS}_{{\bf u}_b}$, i.e. the same matrix  appearing in the Newton computation of the base flow, and  $B$ is a ``weight" matrix associated to the scalar product $\left<v,u\right> = \int \overline{\bf v} \cdot {\bf u} \mbox{ d} \Omega$.

@@

When it comes to the numerical resolution of generalized eigenvalue problems such as $A X = \lambda B X$ (or its adjoint version (\ref{Eigen_Adj_matricial}) ), several methods are possible. Direct methods to compute the whole spectrum are both costly prohibitive and useless. A popular alternative is the use of iterative methods which allow us to compute a limited set of eigenvalues located in the vicinity of a 
``shift" value $\lambda_{shift}$. The simplest version of this method is the simple shift-invert iteration, which consists of solving iteratively the system
$$
X^{n} =  (A- \lambda_{shift} B)^{-1} B X^{n-1}.
$$ 
It is easy to show that this iterative procedure quickly asymptotes to $X^{n+1} \approx ({\lambda^*}^{-1})^n \hat{X}$
where $\hat{X}$ is the eigenmode with largest ${\lambda^*}^{-1}$ (i.e. the one with eigenvalue $\lambda$ closest to the shift).

When a good estimation of the eigenvalue is available, this method converges very rapidly and is very efficient, but it can only provide a single eigenvalue.
If we want to compute a larger number of eigenvalues, we can revert to a generalized version of iterative methods, called Arnoldi methods\cite{Arnoldi51}. The shift-invert version of the Arnoldi method is in fact the most commonly used method of the current time and is at the basis of both the popular matlab function {\tt{eigs}} and the eigenvalue solver of FreeFem (i.e. ARPACK++). Our implementation in StabFem allows to chose between single eigenvalue computation (power method) and multiple eigenvalue computation (Arnoldi). The selection is made according to the parameter ``nev" transmitted to the driver.

@@

As for any numerical method, a crucial point in the numerical efficiency is the design of the mesh. The finite element method allows to use unstructured mesh and hence to locally adapt the refinement. %Following Sipp \& Lebedev, 
The most common procedure is to decompose the domain into several parts with different grid densities; for instance for the wake of a cylinder, we will design a near-wall region with very small  size, a ``wake" region with intermediate mesh size, and an outer region with large mesh size. The inconvenient is that the design relies on an a priori expectation of the regions where gradients will be large.

@@

Let's consider now the matlab code reported in figure \ref{Listing2}. First we build an initial mesh (line 1), and compute base flow solutions for increasing values of the Reynolds number up to $Re = 60$ (lines 3-9).
Then we perform the mesh adaptation with the structural sensitivity, as explained in a previously (line 13). 
The resulting mesh, depicted in figure \ref{fig:mesh2}, is used for the rest of the computations presented in this paper (except for displaying the eigenmode in figure 4$a$). 
Appendix D presents additional test regarding mesh convergence, and demonstrate that results obtained with the resulting mesh are trustable within $0.1\%$ accuracy for the eigenvalue, while requiring a very reasonable number of grid points compared to previous studies of this problem.

@@

In the past decade, efforts have been devoted to extend the range of validity of global approaches into the nonlinear regime for $Re>Re_c$, with the double objective to describe the properties of the limit cycle reached after saturation and to derive amplitude equations describing the transitent dynamics towards this cycle. The two main milestones in this direction  are the {\em weakly nonlinear model}  (WNL) of Sipp \& Lebedev \cite{SippLebedev} and the {\em self-consistent model} (SC) of Manti\v{c}-Lugo et al \cite{MLugo2014}.
 In this review, we will address only the first of the two questions, namely  the description of the saturated cycle, and leave aside the question of transient dynamics. We will successively review the two aforementioned models, in a simplified formulation devoted to describe only the saturated cycle. We also provide a simple implementation of these two models. Finally, in line with the previous section, we will again illustrate with results obtained for the wake of a cylinder.

@@

 In this part, to simplify the notations, we symbolically write the Navier-Stokes equations, as $\partial_t {\bf u} = {\cal NS} ({\bf u})$, therefore dropping the systematic reference to the incompressibility constraint and associated pressure field. The same is done with the linearized operator ${\cal LNS}_{{\bf U}}( {\bf u})$.

@@

\paragraph{WNL model: analysis}

@@

We first review the weakly nonlinear model of Sipp \& 
Lebedev\cite{SippLebedev}, also discussed by Gallaire et al. \cite{FDR2016}.  
The initial derivation of \cite{SippLebedev} makes use of a multiple scale method in order to obtained an amplitude equation.
This complete analysis is reproduced in appendix B. In the present paragraph, we give a simplified derivation of this model restricted to the description of the periodic saturated cycle.
The starting point can be taken as the following expansion of the velocity flow field:
\be{WNL1}
\begin{aligned}
&{\bf u} = {}  {\bf u}_{bc} + \epsilon \left[ A_{wnl}  \hat{\bf u} e^{i (\omega_c+\epsilon^2 \omega_\epsilon)  t} + c.c. \right]\\
&+ \epsilon^2 \left[ {\bf u}_\epsilon + |A_{wnl}|^2  {\bf u}_{2,0} + \left(  A_{wnl}^2 {\bf u}_{2,2} e^{2 i(\omega_c+\epsilon^2 \omega_\epsilon)  t} + c.c. \right) \right]\\
& + {\cal O}( \epsilon ^3),
\end{aligned}
\ee

This expansion is built as an asymptotic expansion in terms of the small parameter $\epsilon = \sqrt{1/Re_c - 1/Re}$ corresponding to the distance to the critical Reynolds. 
The zero-order term ${\bf u}_{bc}$ is the base flow at the threshold $Re_c$.

@@

The second-order term contains three contributions: ${\bf u}_{\epsilon}$ is the modification to the base flow related to the 
increase of $Re$, ${\bf u}_{2,0}$  represents the nonlinear interaction of $\hat{\bf u}$ with its conjugate, ${\bf u}_{2,2}$ the nonlinear 
interaction of $\hat{\bf u}$ with itself. These three terms are computed as solutions of nonsingular linear systems 
(see eqs. \ref{eq:LL1}, \ref{eq:LL2}, \ref{eq:LL3} in appendix B).

At ${\cal O}( \epsilon ^3)$, compatibility conditions have to be imposed to ensure that the problem is correctly posed. These conditions lead to an ``amplitude equation" which relates the amplitude $ A_{wnl}$ to three parameters $\Lambda, \nu_{0}$ and $\nu_{2}$ which depend
uniquely on ${\bf u}_\epsilon$, ${\bf u}_{2,0}$ and ${\bf u}_{2,2}$, respectively (see equations \ref{LAMBDA}, \ref{NU0} and \ref{NU2} in appendix B). Restricting to the description of the limit cycle, the ``amplitude equation" takes the form :

@@

Finally, as specified above, we explain how the mean drag $F_{x,0}$ and fundamental components $(F_{y,1,c},F_{y,1,s})$ of the oscillating lift can be predicted from the WNL approach.
The mean drag  can be obtained by $F_{x,0} = {\cal D}_{Re}({\bf u}_m,p_m)$, where ${\cal D}$ is the drag operator defined in \ref{drag} and 
$[{\bf u}_m,p_m]$ is the {\em mean flow}  
which corresponds to the time-average of expansion  \ref{WNL1}, namely :
$
[{\bf u}_m,p_m] = [{\bf u}_{bc},p_{bc}] +\epsilon^2 \left( [{\bf u}_\epsilon,p_\epsilon] + A_{wnl}^2 [{\bf u}_{2,0}, p_{2,0} ] \right).
$
Developing the terms as an asymptotic expansion leads to :
\be{drag_cos}
 F_{x,0}(Re) \approx F_{x,0,{Re_c}} +  F_{x,0,\epsilon} \left(\frac{1}{Re_c}-\frac{1}{Re}\right)
\ee 
\begin{equation*}
\begin{aligned}
& \mbox{with} \quad F_{x0,{Re_c}}  = {\cal D}_{Re_c} ({\bf u}_{bc},p_{bc}), \quad \mbox{and}  
\\ 
& F_{x0,\epsilon} = {\cal D}_{Re_c} ({\bf u}_\epsilon, p_\epsilon ) - {\cal D}_{1} ({\bf u}_{bc},0) 
+ | A_{wnl} |^2 {\cal D}_{Re_c} ({\bf u}_{2,0},p_{2,0}). 
\end{aligned}
\end{equation*}
%_{\epsilon},p_\epsilon)
%D({\bf u}_{\epsilon})- 2\int_{\Gamma_{cyl}}} \mathbf{D} ({\bf u}_{bc}) \cdot {\bf n}  \cdot {\bf e}_x d \ell \right) 
%\\
%&+|A|^2 D({\bf u}_{2,0})\\
%&+2|A|^2\big(F_{x,r}({\bf u}_{2,2})\cos(2\omega t)
%-F_{x,i}({\bf u}_{2,2})\sin(2\omega t) \big),
